Variants of G\"{o}del's Ontological Proof \\ in a Natural Deduction Calculus

### 2016-04-06 Papers/InProgress/NaturalDeduction/GodProof-ND.tex
Comment out ToDo.
Add missing citation
Make minor improvements in the text

In addition to the five axioms and three definitions that pertain specifically to the argument, the proof also uses {\bf S5}'s iteration principle (in the proof of Lemma 2) and the equality axiom of reflexivity (in the proof of Theorem 1). The use of reflexivity could be avoided, but we chose to use it because it allows the construction of a shorter proof.

@@

In this section we present a new proof in {\bf KB}. This proof uses exactly the same axioms and definitions of Scott's proof shown in the previous sections, but it uses neither the equality axiom of reflexivity nor {\bf S5}'s modal iteration principle. Instead, it relies only on Brouwer's reduction principle (for Theorem 3). The new proof is also shorter than Scott's proof.

@@

In G\"odel's ontological proof, we are proving a restricted modal collapse, which applies to one specific formula, the existence of a God-like being. The interest in the proof naturally decreases if a consequence of the axiomatization is a modal collapse for \emph{all} formulas. Therefore, an improvement would be obtained if the modal collapse was limited only to the property of being god-like or at least to a restricted collection of properties. Several solutions to the problem of modal collapse have been proposed.
Fitting, for instance, has argued that greater care has to be taken with the semantics of higher-order modal logics. Quantified variables may be rigid or flexible; and properties may be treated as intensional or extensional. Making the right choices may prevent the modal collapse \citep{fitting}[Sections 11.9 and 11.10]. Koon \citep{koon}, Anderson \citep{and90}[p. 292] and Sobel \citep{sobel2}[p. 133] also discuss the idea that the notion of property over which quantification is allowed might be too general and restrictions might be appropriate. Two alternative and more conservative solutions, requiring only modifications of the axioms and definitions but no change in the logic, are discussed in greater detail in the next sections.

@@

\begin{theorem}
\label{A:T1}
Positive properties are possibly exemplified:
$$
\all \varphi. [P(\varphi) \imp \pos \ex x.\varphi(x)]
$$
\end{theorem}
\begin{proof}
This proof is omitted, because any of the proofs of Theorem 1 in Sections \ref{sec:scott} and \ref{sec:newproof} would suffice to prove this theorem with only small modifications: the equivalence of axiom \ref{A1} would not need to be reduced to an implication by conjunction elimination. 
\end{proof}

@@

Bj{\o}rdal's variant \citep{bjordal,fuhrmann} is another solution to the Modal Collapse. In striking contrast to other variants, Bj\o rdal's axiomatization takes god-likeness ($G_B$) as a primitive notion, upon which the notion of positive property is defined. This variant also significantly reduces the number of necessary axioms. Scott's variant of Axioms 2, 3 and 4 become unnecessary to prove the final theorem from Bj{\o}rdal's modified definitions 1, 2 and 3 and axioms 1 and 5.

@@

Bj{\o}rdal refines the notion of essence (which he calls maximal composite) by restricting the attention to positive properties only. He calls this refined notion of essence a maximal composite.

@@

ToDo: Fix overfull hboxes

@@

>>>>

In addition to the five axioms and three definitions that pertain specifically to the argument, the proof also uses {\bf S5}'s iteration principle (in the proof of Lemma 2) and the axiom of reflexivity for equality (in the proof of Theorem 1). The use of reflexivity could be avoided, but we chose to use it because it allows the construction of a shorter proof.

@@

In this section we present a new proof in {\bf KB}. This proof uses exactly the same axioms and definitions of Scott's proof shown in the previous sections, but it uses neither the axiom of reflexivity for equality nor {\bf S5}'s modal iteration principle. Instead, it relies only on Brouwer's reduction principle (for Theorem 3). The new proof is also shorter than Scott's proof.

@@

In G\"odel's ontological proof, we are proving a restricted modal collapse, which applies to one specific formula, the existence of a God-like being. The interest in the proof naturally decreases if a consequence of the axiomatization is a modal collapse for \emph{all} formulas. Therefore, an improvement would be obtained if the modal collapse was limited only to the property of being god-like or at least to a restricted collection of properties. Several solutions to the problem of modal collapse have been proposed.
Fitting, for instance, has argued that greater care has to be taken with the semantics of higher-order modal logics. Quantified variables may be rigid or flexible; and properties may be treated as intensional or extensional. Making the right choices may prevent the modal collapse \citep{fitting}[Sections 11.9 and 11.10]. Koon \citep{koons}, Anderson \citep{and90}[p. 292] and Sobel \citep{sobel2}[p. 133] also discuss the idea that the notion of property over which quantification is allowed might be too general and restrictions might be appropriate. Two alternative and more conservative solutions, requiring only modifications of the axioms and definitions but no change in the logic, are discussed in greater detail in the next sections.

@@

\begin{theorem}
\label{A:T1}
Positive properties are possibly exemplified:
$$
\all \varphi. [P(\varphi) \imp \pos \ex x.\varphi(x)]
$$
\end{theorem}
\begin{proof}
This proof is omitted, because any of the proofs of Theorem 1 in Sections \ref{sec:scott} and \ref{sec:newproof} would suffice to prove this theorem with only a small modification: the equivalence of axiom \ref{A1} would not need to be reduced to an implication by conjunction elimination. 
\end{proof}

@@

Bj{\o}rdal's variant \citep{bjordal,fuhrmann} is another solution to the Modal Collapse. In striking contrast to other variants, Bj\o rdal's axiomatization takes god-likeness ($G_B$) as a primitive notion, upon which the notion of positive property is defined. This variant also significantly reduces the number of necessary axioms. Scott's variants of Axioms 2, 3 and 4 become unnecessary to prove the final theorem from Bj{\o}rdal's modified definitions 1, 2 and 3 and axioms 1 and 5.

@@

Bj{\o}rdal refines the notion of essence by restricting the attention to positive properties only. He calls this refined notion of essence a maximal composite.

@@

%ToDo: Fix overfull hboxes

@@

\bibitem[{\itshape Koons(2006)}]{koons}
Koons, R. 2006. Sobel on G\"odel's Ontological Proof. {\itshape Philosophia Christi}, 2:235-248.
